---
title: "EDA"
format: html
editor: visual
---

## Name: Xingwang Yu

## Final project for ST558

## Introduction

briefly describes the data and the variables you have to work with (just discuss the ones you’ll investigate/use in your analysis).

• describes the purpose of your EDA and ultimate goal of modeling.

diabetes \_ binary \_ health \_ indicators \_ BRFSS2015.csv is a clean dataset of 253,680 survey responses to the CDC's BRFSS2015. The target variable Diabetes_binary has 2 classes. 0 is for no diabetes, and 1 is for prediabetes or diabetes. This dataset has 21 feature variables and is not balanced.

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

## Data

```{r setup, warning=FALSE}
library(tidyverse)
library(tidyr)
library(caret)
library(rpart)
library(randomForest)
library(gbm)
```

First, we read in the data and check the variables via `str()`.

```{r}
df <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")
str(df)
```

-   We can note that all the variables are category variables except `BMI`, which is the only one numeric variable.

-   We need to covert the category variables to factors with meaningful names.

```{r}
df <- df |>
  mutate(Diabetes_binary = factor(Diabetes_binary,
                                  levels = c(0, 1, 2),
                                  labels = c("no diabetes", "prediabetes", "diabetes")))
```

The `echo: false` option disables the printing of code (only output is displayed).

# [Click here for the Modeling Page](Modeling.html)
